Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-15 analysis/paper/paper.Rmd
IMPROVING FIG CAPTIONS + ODDS & ENDS

```{r setup, include = FALSE}
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a
movement model to predict locations at specific times. However, the sparse nature of data
from some tags, especially those mounted to the rear flipper, resulted in poor modeling
performance or convergence issues. For this study, we calculated a weighted average daily
location where the inverse of the estimated Argos location error was used for the weight.
Any days where haul-out observations were present without any location data were filled
from the last calculated daily location, and any days where the location intersected with
land were removed from the dataset. Figure \@ref(fig:dataMap) shows the spatial
distribution of available haul-out behavior data for each species across the study area.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
[@teneyck2017]. After fitting GLMPM models, we instead used "type III" F-tests to
calculate p-values [@verhoef2010cs] to evaluate model performance and important terms. We
also produced predictions of haul-out behavior as a function of influential predictors
(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
based on daily or hourly smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces, explore marginal
effects of weather covariates, and determine annual peaks in haul-out activity. The timing
of peak haul-out behavior was further used to regress against the annual maximum sea ice
extent in the study region.

@@

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

@@

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Predicted haul-out probability of ribbon seals across a range of weather conditions."}

@@

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("solar hour") +
  ggtitle("Spotted seals rest on sea ice in late spring and around solar noon.")
p
```

@@

```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Predicted haul-out probability of spotted seals across a range of weather conditions."}
library(ggsci)

@@

p <- ggplot(plot_df,aes(day,solar_hour,fill = ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("solar hour") +
  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
p
```

@@

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Predicted haul-out probability of bearded seals across a range of weather conditions."}

@@

```{r maxextentHO, fig.height = 5.5, include=TRUE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area. While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
ggplot(data = peak_ho_df, aes(y = yday, x = max_extent_km2, color = species)) + 
  geom_smooth(method = "lm", alpha = 0.2, aes(fill = species)) +
  geom_point() +
  facet_grid(age_sex~.) +
  scale_y_continuous(breaks = c(125,135,145, 156),
                     labels = c("05-May", "15-May", "25-May", "05-Jun")) +
  scale_color_manual(values = c('#E66100','#5D3A9B'), name = "") +
  scale_fill_manual(values = c('#E66100','#5D3A9B'), name = "") +
  theme(legend.position = "top",
        legend.justification = "right") +
  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km^2 / 1000)",
       title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited")

>>>>

```{r setup, include = FALSE}
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a
movement model to predict locations at specific times. However, the sparse nature of data
from some tags, especially those mounted to the rear flipper, resulted in poor modeling
performance or convergence issues. For this study, we calculated a weighted average daily
location where the inverse of the estimated Argos location error was used for the weight.
Any days where haul-out observations were present without any location data were filled
from the last calculated weighted average daily location, and any days where the location
intersected with land were removed from the dataset. Figure \@ref(fig:dataMap) shows the
spatial distribution of available haul-out behavior data for each species across the study
area.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
[@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms. We
also produced predictions of haul-out behavior as a function of influential predictors
(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
based on daily or hourly smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces, explore marginal
effects of weather covariates, and determine annual peaks in haul-out activity. The timing
of peak haul-out behavior was further used to regress against the annual maximum sea ice
extent in the study region.

@@

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

@@

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Variability in predicted haul-out probability of ribbon seals across a range of weather conditions thea range of weather conditions encountered in the observed data. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability. Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}

@@

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Spotted seals rest on sea ice in late spring and around solar noon.")
p
```

@@

```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Variability in the predicted haul-out probability of spotted seals across a range of weather conditions thea range of weather conditions encountered in the observed data. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability. Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}
library(ggsci)

@@

p <- ggplot(plot_df,aes(day,solar_hour,fill = ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Bearded seals rest on sea ice around solar noon and solar midnight")
p
```

@@

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across a range of weather conditions thea range of weather conditions encountered in the observed data. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability. Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}

@@

```{r maxextentHO, fig.height = 5.5, include=TRUE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area. While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}
ggplot(data = peak_ho_df, aes(y = yday, x = max_extent_km2, color = species)) + 
  geom_smooth(method = "lm", alpha = 0.2, aes(fill = species)) +
  geom_point() +
  facet_grid(age_sex~.) +
  scale_y_continuous(breaks = c(125,135,145, 156),
                     labels = c("05-May", "15-May", "25-May", "05-Jun")) +
  scale_color_manual(values = c('#E66100','#5D3A9B'), name = "") +
  scale_fill_manual(values = c('#E66100','#5D3A9B'), name = "") +
  theme(legend.position = "top",
        legend.justification = "right") +
  labs(y = "date of peak haul-out", x = "maximum sea ice extent (km^2^ / 1000)",
       title = "The connection between yearly maximum sea ice extent and the \ntiming of peak haul-out probability is limited") +
  theme(axis.text.x = element_markdown())
